Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
slight updates

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
bibliography: bibliography.bib
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author:
  - name: Taimur Khan
    orcid: '0000-0001-7833-5474'
    corresponding: true 
    affiliations:
      - name: Helmholtz Centre for Environmental Research (UFZ)
        address: Theodor-Lieser-Str. 4
        city: Halle (Saale)
        postal-code: 04109 
        country: Germany
        url: taimur.khan@ufz.de
        id: south
    url: https://www.ufz.de/index.php?en=49404
  - name: Jasmin Krebs
    orcid: '0009-0003-9330-1231'
    affiliations:
      - name: Leipzig University
        address: Augustusplatz 10 
        city: Leipzig
        country: Germany
        postal-code: 04109 
        url: jk21byxu@studserv.uni-leipzig.de
  - name: Sharad Kumar Gupta
    orcid: '0000-1111-2222-3333'
    affiliations:
      - ref: south
      - name: Secondary European Affiliation
        address: Tiergartenstr. 17
        city: Heidelberg
        country: Germany
        postal-code: 69121 
        url: sharad.gupta@ufz.de
  - name: Jonathan Renkel
    orcid: '0000-1111-2222-3333'
    affiliations:
      - name: Martin Luther University Halle-Wittenberg
        address: Von-Seckendorff-Platz 2
        city: Halle (Saale)
        country: Germany
        postal-code: 06120
        url: jonathan.renkel@geo.uni-halle.de
  - name: Caroline Arnold
    orcid: '0000-0002-9458-1517'
    affiliations:
      - name: Helmholtz-Zentrum Hereon
        address: Max-Planck-Straße 1
        city: Geesthacht
        country: Germany
        postal-code: 21502 
        url: caroline.arnold@hereon.de
  - name: Nils Nölke 
    orcid: '0000-0003-4925-2287'
    affiliations:
      - name: University of Göttingen, Forest Inventory and Remote Sensing
        address: Büsgenweg 5
        city: Göttingen
        country: Germany
        postal-code: 37077 
        url: nils.noelke@forst.uni-goettingen.de

author-running: T. Khan et al.
abstract: |
  Recent advancements in deep learning have significantly enhanced the extraction of detailed ecological information from multispectral (RGBi) aerial imagery, notably the identification and segmentation of individual tree crowns. Using data from our extensive case study mapping approximately 218.7 million trees across the German federal states of Sachsen and Sachsen-Anhalt, we illustrate how scaling these models beyond controlled settings introduces critical validation challenges. Even minor inaccuracies in tree crown segmentation can substantially impact practical applications in forestry management and urban planning. Our findings emphasize the influence of tree allometry, seasonal variability, shadow effects, and the characteristics of training dataset annotations on model accuracy. Addressing these issues necessitates careful selection and design of training data and exploration of task-specific pre-training through Foundation Models. We highlight these validation complexities and recommend for rigorous steps to ensure the reliability and practical utility of large-scale deep learning models in ecological and urban management contexts.
keywords: [Deep Learning, Ecology, Forestry, Remote Sensing.]
---

@@

To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).

Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges. The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations. Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3). To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).

![a. Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
b. Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for all segmented trees. A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns. This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.]("./figures/figure1.png")

@@

A further fundamental challenge lies in the scarcity of accurate ground-truth data at scale. Reliable validation hinges on high-quality reference data (the "ground truth"), yet collecting detailed crown delineations over large regions is logistically difficult and expensive. For instance, field surveys can map individual trees with great precision, providing highly accurate tree position information (e.g., through GPS measurements of trunk location or detailed canopy spread on the ground) [@steier2024]. However, doing this over thousands of square kilometers is infeasible. Conversely, while aerial and satellite imagery offer broad coverage, even at high resolutions, they present significant challenges for unequivocally labeling each tree crown for validation purposes. This is due to factors like canopy overlap, varied lighting conditions, complex tree morphologies, and the sheer scale of the areas to be annotated, which introduce ambiguities and make consistent manual delineation impractical or prohibitively expensive across large regions. UAV (drone) campaigns can bridge the gap by capturing very high-resolution images or LiDAR of sample areas, but they are limited in flight range and still require extensive human annotation to turn imagery into usable ground truth. The net result is a mismatch of scales: our models aspire to map every tree across entire countries, but our ground truth typically covers only small plots or scattered samples \[^\]. This mismatch means that validating a “wall-to-wall” tree map often involves extrapolating from a tiny fraction of ground-referenced trees, introducing uncertainty. Moreover, ground-reference datasets may not capture the full diversity of conditions (species, canopy shapes, management regimes, etc.) present in the larger mapping area, biasing the validation. Expanding ground-truth collection – through automated methods or crowdsourcing – is thus not just a recommendation but a necessity to overcome this validation bottleneck (as we discuss later).

@@

![Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.]("./figures/figure2.png")

>>>>

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
bibliography: bibliography.bib
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author:
  - name: Taimur Khan
    orcid: '0000-0001-7833-5474'
    corresponding: true 
    affiliations:
      - name: Helmholtz Centre for Environmental Research (UFZ)
        address: Theodor-Lieser-Str. 4
        city: Halle (Saale)
        postal-code: 04109 
        country: Germany
        url: taimur.khan@ufz.de
        id: south
    url: https://www.ufz.de/index.php?en=49404
  - name: Jasmin Krebs
    orcid: '0009-0003-9330-1231'
    affiliations:
      - name: Leipzig University
        address: Augustusplatz 10 
        city: Leipzig
        country: Germany
        postal-code: 04109 
        url: jassi.krebs@gmx.de
  - name: Sharad Kumar Gupta
    orcid: '0000-1111-2222-3333'
    affiliations:
      - ref: south
      - name: Secondary European Affiliation
        address: Tiergartenstr. 17
        city: Heidelberg
        country: Germany
        postal-code: 69121 
        url: sharad.gupta@ufz.de
  - name: Jonathan Renkel
    orcid: '0000-1111-2222-3333'
    affiliations:
      - name: Martin Luther University Halle-Wittenberg
        address: Von-Seckendorff-Platz 2
        city: Halle (Saale)
        country: Germany
        postal-code: 06120
        url: jonathan.renkel@geo.uni-halle.de
  - name: Caroline Arnold
    orcid: '0000-0002-9458-1517'
    affiliations:
      - name: Helmholtz-Zentrum Hereon
        address: Max-Planck-Straße 1
        city: Geesthacht
        country: Germany
        postal-code: 21502 
        url: caroline.arnold@hereon.de
  - name: Nils Nölke 
    orcid: '0000-0003-4925-2287'
    affiliations:
      - name: University of Göttingen, Forest Inventory and Remote Sensing
        address: Büsgenweg 5
        city: Göttingen
        country: Germany
        postal-code: 37077 
        url: nils.noelke@forst.uni-goettingen.de

author-running: T. Khan et al.
abstract: |
  Recent advancements in deep learning have significantly enhanced the extraction of detailed ecological information from multispectral (RGBi) aerial imagery, notably the identification and segmentation of individual tree crowns. Using data from our extensive case study mapping approximately 218.7 million trees across the German federal states of Sachsen and Sachsen-Anhalt, we illustrate how scaling these models beyond controlled settings introduces critical validation challenges. Even minor inaccuracies in tree crown segmentation can substantially impact some practical applications in forestry management and urban planning. Our findings emphasize the influence of tree allometry, seasonal variability, shadow effects, and the characteristics of training dataset annotations on model accuracy. Addressing these issues necessitates careful selection and design of training data and exploration of task-specific pre-training through Foundation Models. We highlight these validation complexities and recommend for rigorous steps to ensure the reliability and practical utility of large-scale deep learning models in ecological and urban management contexts.
keywords: [Deep Learning, Ecology, Forestry, Remote Sensing.]
---

@@

To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). <!--The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).-->

Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges. The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations. Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3). To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).

![a. Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns. b. Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for all segmented trees. A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns. This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.](./figures/figure1.png)

@@

A further fundamental challenge lies in the scarcity of accurate ground-truth data at scale. Reliable validation hinges on high-quality reference data (the "ground truth"), yet collecting detailed crown delineations over large regions is logistically difficult and expensive. For instance, field surveys can map individual trees with great precision, providing highly accurate tree position information (e.g., through GPS measurements of trunk location or detailed canopy spread on the ground) [@steier2024]. However, doing this over thousands of square kilometers is infeasible. Conversely, while aerial and satellite imagery offer broad coverage, even at high resolutions, they present significant challenges for unequivocally labeling each tree crown for validation purposes. This is due to factors like canopy overlap, varied lighting conditions, complex tree morphologies, and the sheer scale of the areas to be annotated, which introduce ambiguities and make consistent manual delineation impractical or prohibitively expensive across large regions. UAV (drone) campaigns can bridge the gap by capturing very high-resolution images or LiDAR of sample areas, but they are limited in flight range and still require extensive human annotation to turn imagery into usable ground truth. The net result is a mismatch of scales: our models aspire to map every tree across entire countries, but our ground truth typically covers only small plots or scattered samples \[\^\]. This mismatch means that validating a “wall-to-wall” tree map often involves extrapolating from a tiny fraction of ground-referenced trees, introducing uncertainty. Moreover, ground-reference datasets may not capture the full diversity of conditions (species, canopy shapes, management regimes, etc.) present in the larger mapping area, biasing the validation. Expanding ground-truth collection – through automated methods or crowdsourcing – is thus not just a recommendation but a necessity to overcome this validation bottleneck (as we discuss later).

@@

![Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.](%22./figures/figure2.png%22)
